gij bent 2

### 2025-09-28 paper/paper.qmd
Punctuation and text flow

One of the key ideas from usage-based linguistics is that language is not a system that exists just "by itself". Rather, it is widely accepted within this branch that language arises out of the many interactions that language users have with each other [@keller_language_1994]. As such, language becomes a "complex-adaptive system" [@steels_language_2000;@beckner_language_2009;@ellis_what_2012;@kretzschmar_language_2015], a system without a single authoritative control, constructed through the simple yet numerous interactions of its components, language users. While each speaker of a language (be it Dutch, English or Maori) has their own personalised representation of the rules and constraints that they think make up that language, these representations have enough in common in order to warrant talking about a shared "code", or more broadly, a language.

@@

It is essential to recognize that the widespread changes observed across entire populations ultimately stem from changes that begin within individuals. While linguistic changes may eventually become collective phenomena, their origins lie in the minds and speech of individual speakers. This is why both from a usage-based and a cognitive-linguistic perspective the individual should be the primary unit of linguistic analysis where possible. Recent studies into individual variation in language use reveal that linguistic behavior is not uniformly progressive or conservative, either within individuals or among groups [@nevalainen_diffusion_2011]. While later-born authors often show more progressive tendencies, there is notable individual variation even within generational cohorts [@petre_real-time_2018]. There is also variation in the language-internal factors influencing language change. While most language-internal factors are broadly shared, less influential factors reveal personal differences in usage [@fonteyn_individuality_2020]. Moreover, some individuals are more adaptable to linguistic change over their lifetimes, while others remain stable or even deviate in the opposite direction [@fonteyn_probability_2022]. Personality traits also play a role, affecting how individuals adopt or evaluate new linguistic features [@guy_cognitive_2013;@zenner_individual_2016;@van_den_stock_individual_2024].

@@

Linguistically, the effects of networks have been studied before. @milroy_linguistic_1985, who base themselves on @granovetter_strength_1973 and @rogers_communication_1971, demonstrated that in a network both weak and strong ties are essential for the propagation of linguistic change. Strong ties (e.g. close friends and family) maintain and reinforce linguistic norms within tight-knit groups, resisting linguistic innovation. In contrast, weak ties (e.g. acquaintances) form bridges between groups, facilitate the spread of linguistic innovations. Weak ties are especially crucial for exposing individuals to linguistic features or patterns that are not present within their immediate, tightly connected group, enabling the flow of new linguistic forms across disparate groups, broadening the reach of linguistic innovations. Once exposed to these new forms, individuals connected through weak ties may adopt them and, through their strong ties, introduce them to their close-knit circles. This dynamic does not affect all individuals equally: some are positioned to introduce innovations into the network, while others play a crucial role in amplifying and spreading them.

Within the propagation of innovations, different roles can be distinguished. Innovators, who are less central and only loosely connected to others through weak ties, acquiring novel ideas from outside their immediate community [@rogers_diffusion_2003: 282-283]. Early adopters, by contrast, are more centrally positioned and enjoy a higher degree of connectivity within the population. Through their many strong ties, they amplify and legitimize the innovation, spreading it efficiently across their dense local networks [@rogers_diffusion_2003: 282-283]. In this way, innovators and early adopters together bridge the gap between the periphery and the core of a community, ensuring that new linguistic forms do not remain isolated but instead gain broader social traction [@granovetter_strength_1973;@milroy_linguistic_1985]. The idea that innovation originates in the periphery of a network before it can spread has been attested before in network studies [@van_de_leemput_transformation_2024], but not whithin linguistics.

Except for some (often small-scale) studies [e.g. @labov_study_1968;@milroy_belfast_1978;@nevalainen_social_2015;@sharma_scalar_2017;@paradowski_how_2022] and simulations [@ke_language_2008;@ogura_dynamic_2008], it has been near impossible to get an elaborate insight into the social relationships of people on a large scale, let alone how these relationships relate to language. Again, social media offer an opportunity here. While it is not practical to get a full network of *all* social media users, we *can* get information on how "connected" a particular user is. Through metrics like "followers" and "following", we know how many connections social media users have. This allows us to get an idea of the centrality of a user in the larger social media network, even if we do not know the full network itself. This brings us to the second research question in this study:

@@

Dutch as it is spoken in Belgium has evolved strongly since the second half of the 20^th^ century. In the largest part of the 20^th^ century, the language situation of Dutch in Flanders, the Dutch-speaking part of Belgium, could best be described as a *diglossia* [@auer_europe_2005]. This means that the Dutch was characterised by a dichotomy between standard language on the one hand, and dialects on the other hand. These two were the language modes available to speakers of Dutch. A large-scale language policy campaign at the timed attempted to teach those speakers who mastered only their local dialect the standard language. While these efforts were well intended, the imposition of a "foreign" standard language -- the standard was imported from the Netherlands -- were met with limited enthusiasm by language users in Flanders. However, this language campaign was not without effect, albeit not the intended one.

@@

To estimate the individual variation among our tweet authors, we would ideally have a quantitative measure which indicates the deviation of each user from the expected linguistic behaviour. Authors who do not deviate would be assigned a (near-)zero value, while those who *do* deviate would receive either a positive or negative value, depending on whether they exceed the expectations for their profiles, or actually go against it. In this way, our measure of individual variation truly expresses the peculiarities of each individual speaker.

@@

As mentioned in @sec-introduction, the networked aspect of social media data provides an opportunity for researchers to investigate which people use what variant, and what that says about the diffusion of a new variant. Since our dataset comes from the Twitter platform, we also dispose of metrics which define a user's relative popularity on the platform. The first metric, "Followers", tallies how many other users want to see a user's posts appear in their own social media feed. Users with a large number of followers have a broad reach, and therefore have the possibility to be influential in their language use. The second metric, "Following", states how many other people's posts a user wants to see in their social media feed. Users who follow a large number of other users have the opportunity of being exposed to a larger linguistic pallette -- though this need not necessarily be the case.

@@

To answer this question, we built a common linear regression model. We model the coefficients in terms of the influence of the users, but we also supplement our analysis with the dialect and gender distinctions we discussed earlier. Bundling all these possible influences in the same model will ensure that we do not accidentally assign variation between language users to the wrong source. It will also function as an extra check for the results from @sec-group-variation, as we will now be able to check statistically whether the distinctions we spotted in the visual exploration are statistically significant distinctions. @tbl-linear-predictors gives an overview of the predictors that were used in the linear regression model.

@@

Finally, we look at the network influence parameter. Recall that our influence predictor expresses the (im)balance between the "followers" and "following" metrics, with those users who are followed by more people than they follow themselves being attributed a larger influence. The regression table in @tbl-linear-predictors shows that, as our influence metric increases, so does the probability for the use of *gij bent*. This means that the more influential a speaker is in the Twitter network, the more likely they are to use the new, innovative *gij bent* form. In other words, the innovative *gij bent* form is not confined to peripheral innovators but is actively promoted by more central and prestigious figures in the Twitter network. This finding aligns with network-based theories of diffusion, in which highly connected early adopters play a crucial role in amplifying and legitimizing innovations [@rogers_diffusion_2003: 282-283]. It also indicates that *gij bent* may have progressed beyond the initial stage of peripheral innovation, i.e. the stage in which the innovators are crucial, and entered a phase in which socially visible and influential users help to normalize the form.

The observed synchronic variation can also be interpreted through the framework of binary language change, where a new variant, *gij bent*, competes with an older one, *gij zijt*. Such binary changes are typically modeled using an S-curve [@speelman_logistic_2014], which plots the cumulative adoption of the innovation over time. This S-curve has a bell-shaped counterpart, which does not capture the cumulative adoption, but instead captures the rate of adoption at any given point in time: few individuals adopt at the very beginning, adoption peaks as the innovation gains momentum, and then levels off once most speakers have shifted. Diffusion research has shown that the individuals driving this process can be categorized into five adopter groups, defined mathematically by standard deviations around the mean adoption time [@rogers_diffusion_2003: PAGES]: innovators, early adopters, early majority, late majority, and laggards (see FIGURE BELOW). As already hinted at in @sec-introduction, each of these categories is embedded in the network in a different way. Innovators, positioned at the periphery and connected through weak ties, acquire innovations from outside their immediate community and initiate change. Early adopters, in contrast, are more centrally embedded and exert strong influence through their many strong ties, thereby accelerating diffusion and granting legitimacy to the innovation [REFS]. Our finding that more influential Twitter users are more likely to employ *gij bent* suggests that this form has moved beyond the initial innovator stage and is now being amplified by early adopters. In this way, we can infer in which stage of the S-curve the change is situated: between the second and the first standard deviation before the mean.

@@

## Conclusion

>>>>

One of the key ideas from usage-based linguistics is that language is not a system that exists just "by itself". Rather, it is widely accepted within this branch that language arises out of the many interactions that language users have with each other [@keller_language_1994]. As such, language becomes a "complex-adaptive system" [@steels_language_2000;@beckner_language_2009;@ellis_what_2012;@kretzschmar_language_2015], a system without a single authoritative control, constructed through the simple yet numerous interactions of its components: language users. While each speaker of a language (be it Dutch, English or Maori) has their own personalised representation of the rules and constraints that they think make up that language, these representations have enough in common in order to warrant talking about a shared "code", or more broadly, a language.

@@

It is essential to recognize that the widespread changes observed across entire populations ultimately stem from changes that begin within individuals. While linguistic changes may eventually become collective phenomena, their origins lie in the minds and speech of individual speakers. This is why both from a usage-based and a cognitive-linguistic perspective, the individual should be the primary unit of linguistic analysis where possible. Recent studies into individual variation in language use reveal that linguistic behavior is not uniformly progressive or conservative, either within individuals or among groups [@nevalainen_diffusion_2011]. While later-born authors often show more progressive tendencies, there is notable individual variation even within generational cohorts [@petre_real-time_2018]. There is also variation in the language-internal factors influencing language change. While most language-internal factors are broadly shared, less influential factors reveal personal differences in usage [@fonteyn_individuality_2020]. Moreover, some individuals are more adaptable to linguistic change over their lifetimes, while others remain stable or even deviate in the opposite direction [@fonteyn_probability_2022]. Personality traits also play a role, affecting how individuals adopt or evaluate new linguistic features [@guy_cognitive_2013;@zenner_individual_2016;@van_den_stock_individual_2024].

@@

Linguistically, the effects of networks have been studied before. @milroy_linguistic_1985, who base themselves on @granovetter_strength_1973 and @rogers_communication_1971, demonstrated that in a network, both weak and strong ties are essential for the propagation of linguistic change. Strong ties (e.g. close friends and family) maintain and reinforce linguistic norms within tight-knit groups, resisting linguistic innovation. In contrast, weak ties (e.g. acquaintances) form bridges between groups, facilitate the spread of linguistic innovations. Weak ties are especially crucial for exposing individuals to linguistic features or patterns that are not present within their immediate, tightly connected group, enabling the flow of new linguistic forms across disparate groups and broadening the reach of linguistic innovations. Once exposed to these new forms, individuals connected through weak ties may adopt them and, through their strong ties, introduce them to their close-knit circles. This dynamic does not affect all individuals equally: some are positioned to introduce innovations into the network, while others play a crucial role in amplifying and spreading them.

Within the propagation of innovations, different roles can be distinguished. Innovators, who are less central and only loosely connected to others through weak ties, acquire novel ideas from outside their immediate community [@rogers_diffusion_2003: 282-283]. Early adopters, by contrast, are more centrally positioned and enjoy a higher degree of connectivity within the population. Through their many strong ties, they amplify and legitimize the innovation, spreading it efficiently across their dense local networks [@rogers_diffusion_2003: 282-283]. In this way, innovators and early adopters together bridge the gap between the periphery and the core of a community, ensuring that new linguistic forms do not remain isolated but instead gain broader social traction [@granovetter_strength_1973;@milroy_linguistic_1985]. The idea that innovation originates in the periphery of a network before it can spread has been attested before in network studies [@van_de_leemput_transformation_2024], but not whithin linguistics.

Except for some (often small-scale) studies [e.g. @labov_study_1968;@milroy_belfast_1978;@nevalainen_social_2015;@sharma_scalar_2017;@paradowski_how_2022] and simulations [@ke_language_2008;@ogura_dynamic_2008;@fagyal_centers_2010], it has been near impossible to gain an elaborate insight into the social relationships of people on a large scale, let alone how these relationships relate to language. Again, social media offer an opportunity here. While it is not practical to get a full network of *all* social media users, we *can* get information on how "connected" a particular user is. Through metrics like "followers" and "following", we know how many connections social media users have. This allows us to get an idea of the centrality of a user in the larger social media network, even if we do not know the full network itself. This brings us to the second research question in this study:

@@

Dutch as it is spoken in Belgium has evolved strongly since the second half of the 20^th^ century. In the largest part of the 20^th^ century, the language situation of Dutch in Flanders, the Dutch-speaking part of Belgium, could best be described as a *diglossia* [@auer_europe_2005]. This means that Belgian Dutch was characterised by a dichotomy between standard language on the one hand, and dialects on the other hand. These two were the language modes available to speakers of Dutch. A large-scale language policy campaign at the time attempted to teach those speakers who mastered only their local dialect the standard language. While these efforts were well intended, the imposition of a "foreign" standard language -- the standard was imported from the Netherlands -- were met with limited enthusiasm by language users in Flanders. However, this language campaign was not without effect, albeit not the intended one.

@@

To estimate the individual variation among our tweet authors, we would ideally have a quantitative measure which indicates the deviation of each user from the expected linguistic behaviour. Authors who do not deviate would be assigned a (near-)zero value, while those who *do* deviate would receive either a positive or negative value, depending on whether they exceed the expectations for their profiles, or go against it. In this way, our measure of individual variation truly expresses the peculiarities of each individual speaker.

@@

As mentioned in @sec-introduction, the networked aspect of social media data provides an opportunity for researchers to investigate which people use what variant, and what that says about the diffusion of a new variant. Since our dataset comes from the Twitter platform, we also dispose of metrics which define a user's relative popularity on the platform. The first metric, "Followers", tallies how many other users want to see a user's posts appear in their own social media feed. Users with a large number of followers have a broad reach, and therefore have the possibility to be influential in their language use. The second metric, "Following", states how many other people's posts a user wants to see in their social media feed. Users who follow a large number of other users have the opportunity of being exposed to a larger linguistic pallette, though this need not necessarily be the case.

@@

To answer this question, we built a linear regression model. We model the coefficients in terms of the influence of the users, but we also supplement our analysis with the dialect and gender distinctions we discussed earlier. Bundling all these possible influences in the same model will ensure that we do not accidentally assign variation between language users to the wrong source. It will also function as an extra check for the results from @sec-group-variation, as we will now be able to check statistically whether the distinctions we spotted in the visual exploration are statistically significant distinctions. @tbl-linear-predictors gives an overview of the predictors that were used in the linear regression model.

@@

Finally, we look at the network influence parameter. Recall that our influence predictor expresses the (im)balance between the "followers" and "following" metrics, with those users who are followed by more people than they follow themselves being attributed a larger influence. The regression table in @tbl-linear-predictors shows that, as our influence metric increases, so does the probability for the use of *gij bent*. This means that the more influential a speaker is in the Twitter network, the more likely they are to use the new, innovative *gij bent* form. In other words: the innovative *gij bent* form is not confined to peripheral innovators, but is actively promoted by more central and prestigious figures in the Twitter network. This finding aligns with network-based theories of diffusion, in which highly connected early adopters play a crucial role in amplifying and legitimising innovations [@rogers_diffusion_2003: 282-283]. It also indicates that *gij bent* may have progressed beyond the initial stage of peripheral innovation, i.e. the stage in which the innovators are crucial, entering a phase in which socially visible and influential users help to normalize the form.

The observed synchronic variation can also be interpreted through the framework of binary language change, where a new variant, *gij bent*, competes with an older one, *gij zijt*. Such binary changes are typically modeled using an S-curve [@speelman_logistic_2014], which plots the cumulative adoption of the innovation over time. This S-curve has a bell-shaped counterpart, which does not capture the cumulative adoption, but instead captures the rate of adoption at any given point in time: few individuals adopt at the very beginning, adoption peaks as the innovation gains momentum, and then levels off once most speakers have shifted.

Diffusion research has shown that the individuals driving the diffusion process can be categorized into five adopter groups, defined mathematically by standard deviations around the mean adoption time [@rogers_diffusion_2003: PAGES]: innovators, early adopters, early majority, late majority, and laggards (see FIGURE BELOW). As already hinted at in @sec-introduction, each of these categories is embedded in the network in a different way. Innovators, positioned at the periphery and connected through weak ties, acquire innovations from outside their immediate community and initiate change. Early adopters, in contrast, are more centrally embedded and exert strong influence through their many strong ties, thereby accelerating diffusion and granting legitimacy to the innovation [REFS]. Our finding that more influential Twitter users are more likely to employ *gij bent* suggests that this form has moved beyond the initial innovator stage and is now being amplified by early adopters. In this way, we can infer in which stage of the S-curve the change is situated: between the second and the first standard deviation before the mean.

@@

## Discussion and conclusion
